Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-27 trunk/paper/drinking.tex

To illustrate these challenges, consider physiological processes of
the heart. One of the heart's functions is {\em Heart beating}, i.e.,
a capability that is realized through processes of the type {\em Heart
  beating}. {\em Blood} is a participant of {\em Heart beating}
processes.  An abnormal phenotype of an organism could be that the
{\em rate of heart beating} is increased with the intended meaning
that the number of {\em Heart beating} processes in a given time
interval is higher than normal. An attribute such as {\em increased
  rate of heart beating}

Heart rate increased: frequency of CMC increased fluid flow increased
volume per contraction increased ...

what is the relation to the heart?

%%%%%%%%%%%%%%%%%%
observation 1: there is a fundamental distinction between attribution
of qualities for a fixed period (e.g., 24h) and single {\em act}s of
drinking.

observation 2: each of the different possible ``increased'' drinking
behavior may be the result of different underlying causes and due to
this, distinguishing between them may be important.

observation 3: depending on the experiment performed, only some of the
actual qualities can be measured.

observation 4: some qualities can be inferred given others. For
example,...

>>>>

To illustrate these challenges, consider physiological processes of
the heart. One of the heart's functions is {\em Heart beating}, i.e.,
a capability that is realized through processes of the type {\em Heart
  beating}. {\em Blood} is a participant of {\em Heart beating}
processes.  An abnormal phenotype of an organism could be that the
{\em rate of heart beating} is increased. The intended meaning of such
a description is that the number of {\em Heart beating} processes in a
given time interval is higher than normal. Another important attribute
of heart physiology is the fluid flow rate through the heart. For an
abnormal phenotype such as {\em increased rate of fluid flow through
  the heart}, the intended meaning could either be that the amount of
fluid that is moved through the heart within a single heart beating
process is increased or that the amount of fluid that is moved through
the heart within a period of time interval is increased.

Based on these examples, we can make several observations about
process attributes. First, for a process like {\em Heart beating}, we
can distinguish between single occurrences and processes in which {\em
  Heart beating} occurs multiple times. Only the latter kind of
process may have a {\em rate of heart beating}, while {\em Heart
  beating} processes do not have such an attribute. Second, we can
distinguish between abnormal fluid flow rates in {\em Heart beating}
processes and rate of fluid flow through the heart within a given
duration. Both may have entirely different underlying causes and it is
therefore important to distinguish between them. Finally, we may be
able to infer some phenotypes from others, thereby limiting the number
of phenotypes that must be experimentally observed. For example, when
the fluid flow rate in single heart beating processes is increased and
the {\em rate of heart beating} is increased within a process $P$,
then the rate of fluid flow through the heart will be increased for
$P$.

Here, we present the foundations for an ontology of physiological
phenotypes.
